Concurrent Programming in Harmony

### 2022-01-13 book/paper.tex
Improved naming of a file

\noindent
\autoref{fig:queues}(a) shows a sequential specification for such a
queue in Harmony (exploiting some methods from the \texttt{list}
module described in \autoref{ap:list}),
but it cannot be used with threads concurrently accessing this queue.
\autoref{fig:queues}(b) shows the corresponding concurrent specification.
See \autoref{fig:queuedemo} for a simple demonstration program that uses
a concurrent queue.

@@

\begin{figure}
\begin{code}
<{:input:../code/intsettest.hny}>
\end{code}
\caption{\harmonylink{code/intsettest.hny} Test code for set objects}
\label{fig:intsettest}
\end{figure}

@@

\autoref{fig:setobj} gives the specification of a concurrent set object.
\texttt{SetObject()} returns a pointer to a variable that
contains an empty set, rather than returning an empty set \emph{value}.
As such, it is more like an object in an object-oriented language than
like a value in its own right.
Values can be added to the set object using
<{insert()}> or deleted using <{remove()}>.
Method <{contains()}> checks if a particular value is in the list.
\autoref{fig:intsettest} contains a simple (although not very thorough)
test program to demonstrate the use of set objects.

@@

Testing is a way to increase confidence in the correctness
of an implementation.
\autoref{fig:queuedemo} demonstrates how concurrent queues
may be used, but it is not a very thorough test program
for an implementation such as the one in \autoref{fig:queueconc}
and does little to increase our confidence in its correctness.
To wit, if <{get()}> always returned 1, the program
would find no problems.
Similarly, \autoref{fig:intsettest} is not a good test program
for something as complicated as \autoref{fig:linkedlist}.
In this chapter, we will look at approaches to testing
concurrent code.

@@

\begin{figure}
\begin{code}
<{:input:../code/qtest1.hny}>
\end{code}
\caption{\harmonylink{code/qtest1.hny} Concurrent \texttt{put()}}
\label{fig:qtest1}
\end{figure}

If you want a purely black box test for testing a queue implementation,
then adding linearization points is not possible.
We will present a few test
programs that try to identify incorrect behavior of
a black box queue implementation in the presence of concurrency.

\autoref{fig:qtest1} shows a program that
checks concurrent <{put}> operations using
<{N}> threads.  Thread \texttt{putter(<{v}>)}
adds <{v}> to the queue.
Thread \texttt{main()} waits until all \texttt{putter}
threads have finished, and then dequeues <{N}>
items, verifying that they form the set
$\{ 0 .. <{N}> - 1 \}$.

\begin{figure}
\begin{code}
<{:input:../code/qtest2.hny}>
\end{code}
\caption{\harmonylink{code/qtest2.hny} Concurrent <{get()}>}
\label{fig:qtest2}
\end{figure}

Similarly, \autoref{fig:qtest2} shows a program that
checks concurrent <{get}> operations using
<{N}> threads.
The queue is initialized with the sequence
$0 .. <{N}> - 1$.
Thread \texttt{getter(<{v}>)} removes an entry from the
queue and makes sure that no other thread got the
same entry.
Thread \texttt{main()} waits until all <{getter}>
threads have finished and checks to make sure that
the queue is empty.

\begin{figure}
\begin{code}
<{:input:../code/qtest3.hny}>
\end{code}
\caption{\harmonylink{code/qtest3.hny} Mix of concurrent
<{get()}> and \texttt{put()} operations}
\label{fig:qtest3}
\end{figure}

\autoref{fig:qtest3} checks a mixture of
concurrent <{get}> and <{put}> operations.
It starts <{N getter}> threads and
<{N}> \texttt{putter} threads.
Thread <{main}> waits until all have finished,
and then checks to make sure that whatever is left
on the queue was not also consumed by some
<{getter}> thread.
Finally, it makes sure that the queue is empty
once all the values that were added
to it by the \texttt{putter} threads have been consumed.

A common mistake that people make is to use
a globally shared variable in the implementation of
a concurrent data structure.
The <{get()}> method of the queue implementation
uses a local variable <{node}>, declared using
a <{let}> statement, but it would be easy to
mistakenly use a global variable <{node}> instead.
If there is only one queue, this is not an issue.
So, it is important to test not just one queue at a
time, but also multiple queues.
\autoref{fig:qtest4} gives an example that
tests the <{get}> method for use of global
variables.  Even if the assertion does not fail,
Harmony might find a data race if there are global
variables shared between different queues.

While having five different test programs is fine,
it would be nice if we can combine them all in one
test program.
This has to be done with some care, as running the
concurrent test programs simultaneously would
lead to a state space explosion.
Instead, we can leverage the <{choose}>
expression.
\autoref{fig:multitest} shows how one might go
about this.
Harmony will run all the tests, but the
tests do not interfere with one another.

\begin{figure}
\begin{code}
<{:input:../code/qtest4.hny}>
\end{code}
\caption{\harmonylink{code/qtest4.hny} Test with multiple queues}
\label{fig:qtest4}
\end{figure}

\begin{figure}
\begin{code}
<{:input:../code/multitest.hny}>
\end{code}
\caption{\harmonylink{code/multitest.hny} Running a
sequential test followed by several concurrent tests}
\label{fig:multitest}
\end{figure}

@@

\begin{code}
<{:input:../code/baddblwait.hny}>
\end{code}

>>>>

\noindent
\autoref{fig:queues}(a) shows a sequential specification for such a
queue in Harmony (exploiting some methods from the \texttt{list}
module described in \autoref{ap:list}).
It is a credible queue implementation,
but it cannot be used with threads concurrently accessing this queue.
\autoref{fig:queues}(b) shows the corresponding concurrent specification.
It cannot be used as an implementation for a queue, as processors generally
do not have atomic operations on lists, but it will work well as a
specification.
See \autoref{fig:queuedemo} for a simple demonstration program that uses
a concurrent queue.

@@

\begin{figure}
\begin{code}
<{:input:../code/setobjtest.hny}>
\end{code}
\caption{\harmonylink{code/setobjtest.hny} Test code for set objects}
\label{fig:setobjtest}
\end{figure}

@@

\autoref{fig:setobj} gives the specification of a concurrent set object.
\texttt{SetObject()} returns a pointer to a variable that
contains an empty set, rather than returning an empty set \emph{value}.
As such, it is more like an object in an object-oriented language than
like a value in its own right.
Values can be added to the set object using
<{insert()}> or deleted using <{remove()}>.
Method <{contains()}> checks if a particular value is in the list.
\autoref{fig:setobjtest} contains a simple (although not very thorough)
test program to demonstrate the use of set objects.

@@

Testing is a way to increase confidence in the correctness
of an implementation.
\autoref{fig:queuedemo} demonstrates how concurrent queues
may be used, but it is not a very thorough test program
for an implementation such as the one in \autoref{fig:queueconc}
and does little to increase our confidence in its correctness.
To wit, if <{get()}> always returned 1, the program
would find no problems.
Similarly, \autoref{fig:setobjtest} is not a good test program
for something as complicated as \autoref{fig:linkedlist}.
In this chapter, we will look at approaches to testing
concurrent code.

@@

@@

\begin{code}
<{:inline:}>
while forks[left]:
    synch.wait(?conds[left], ?mutex)
while forks[right]:
    synch.wait(?conds[right], ?mutex)
<{:end:}>
\end{code}
